---
title: "Analysis of weather events on human health and economy"
author: "haoeric"
date: "11 July, 2014"
output: html_document
---
     
## Synopsis
Storms and other severe weather events can cause both public health and economic problems 
for communitiesand municipalities. Many severe events can result in fatalities, 
injuries, and property damage, and preventing such outcomes to the extent possible is 
a key concern.


## 1.Data Processing
      
1.1 get the data
Download the file from NOAA Storm Database
```{r dowanload data, cache=TRUE}
#File_Url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
#download.file(file_Url, destfile = "/Users/chenhao/Ranalysis/StormData.csv.bz2", method = "curl")
```
        
1.2 Reading the data
Decompress the file and read it
```{r read the file, cache=TRUE}
#sd <- bzfile("/Users/chenhao/Ranalysis/StormData.csv.bz2", "StormData.csv")
storm_data <- read.csv("/Users/chenhao/Ranalysis/StormData.csv", stringsAsFactors = F)
#unlink(sd)
```
       
1.3 Take a first look at the data
Let's see what's in the dataset
```{r take a glance of the data}
head(storm_data)
str(storm_data)
```

        
## 2.Results
     
2.1 Check all types of weather events:
```{r print weather events and change to factor}
eventype <- sort(unique(storm_data$EVTYPE))
storm_data$EVTYPE <- as.factor(toupper(storm_data$EVTYPE))
```
      
2.2 Plot the top 10 events that cause maximun fatalities and injuries:
```{r plot top harm}
harm <- aggregate(cbind(FATALITIES,INJURIES) ~ EVTYPE, data = storm_data, sum)
harm <- subset(harm, FATALITIES > 0 | INJURIES > 0) 
top_fatalities <- harm[order(-harm$FATALITIES),][1:10,]
top_injuries <- harm[order(-harm$INJURIES),][1:10,]

library(ggplot2)
par(mfrow=c(1,2))
ggplot(data = top_fatalities, aes(EVTYPE, FATALITIES, fill = FATALITIES)) + 
        geom_bar(stat = "identity") + xlab("Weather Event") + ylab("Fatalities") + 
        ggtitle("Fatalities caused by Events (top 10) ") + 
        coord_flip() + theme(legend.position = "none")

ggplot(data = top_injuries, aes(EVTYPE, INJURIES, fill = INJURIES)) + 
        geom_bar(stat = "identity") + xlab("Weather Event") + ylab("Injuries") + 
        ggtitle("Injuries caused by Events (top 10) ") + 
        theme(axis.text.x=element_text(angle = 45, hjust = 1)) + 
        theme(legend.position = "none")
```
       
We can clearly see from the plot that **tornado** are most harmful with respect to population health.
       
2.3 Economic Impact of Weather Events
Take a glance of economic cost related items first
```{r summary cost}
table(storm_data$PROPDMGEXP)
summary(storm_data$PROPDMG)
```
      
2.4 Combine the PROPDMGEX and PROPDMG to get the economic cost caused by different weather events, and see the summary
```{r econimic cost caused by events}
# combine the PROPDMGEX and PROPDMG to get the economic cost
econo_costs <- rep(0, length(storm_data$PROPDMGEX))  

for (i in 1:length(storm_data$PROPDMGEX)){
        if (storm_data$PROPDMGEX[i] == ""){
                econo_costs[i] <- storm_data$PROPDMG[i]
        }
        else{
                unit <- switch(EXPR = storm_data$PROPDMGEX[i],
                               '-' = -1, '?' = 1, '+' = 1, '1' = 1, '2' = 10^2, '3' = 10^3,
                               '4' = 10^4, '5' = 10^5, '6' = 10^6, '7' = 10^7, '8' = 10^8,
                               'h' = 100, 'K' = 1000, 'm' = 10^6, 'B' = 10^9,'0'=1,
                               'H' = 100, 'M' = 10^6
                )
                econo_costs[i] <- storm_data$PROPDMG[i] * unit
        }       
}

storm_data$ECONOMIC_COST <- econo_costs
summary(storm_data$ECONOMIC_COST)
```
       
2.5 Plot the top 10 events that caused the maximum economic cost
```{r plot economic cost}
cost <- aggregate(ECONOMIC_COST ~ EVTYPE, data = storm_data, sum)
# get the top 10 items
top_cost <- cost[order(-cost$ECONOMIC_COST),][1:10,]

ggplot(data = top_cost, aes(EVTYPE, ECONOMIC_COST, fill = ECONOMIC_COST)) + 
        geom_bar(stat = "identity") + xlab("Event") + 
        theme(axis.text.x=element_text(angle = 45, hjust = 1)) + 
        ylab("Economic costs in $") + ggtitle("Economic costs caused by Events (top 10)") + 
        theme(legend.position = "none")
```
        
It's clear that Flood caused the maximum economic cost.


















